### 2024-03-13 outputs/paper/paper.qmd
Updated the data section

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: Table 1
#| fig-cap: Top-ten causes of death in Canada in 2022
#| echo: false

combined_data %>%
  filter(Year == 2021, Ranking <= 10) %>%
  mutate(Number_of_Death = format(Number_of_Death, big.mark = ",")) %>%
  kable(
    col.names = c("Year", "Leading_Cause_Death", "Ranking", "Number_of_Death"),
    align = c("l", "r", "r", "r"),
    digits = 0, booktabs = TRUE, linesep = ""
  )
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-03-13 scripts/02-data_cleaning.R

@@ -14,30 +14,65 @@ library(dplyr)
 ### Clean the data ###
 raw_data <- read.csv("inputs/data/raw_data.csv")
 
-# Select the columns of interest
+# # Select the columns of interest
 col_name <- names(raw_data)
 raw_data <- raw_data %>%
-  rename("Year" = "REF_DATE", 
+  rename("Year" = "REF_DATE",
          "Leading_Cause_Death" = "Leading.causes.of.death..ICD.10.",
          "Value" = "VALUE") %>%
   select("Year", "Characteristics", "Leading_Cause_Death", "Value")
+# 
+# # Subset for Number of Deaths
+# number_data <- filter(raw_data, Characteristics == "Number of deaths") %>%
+#   select(Year, Leading_Cause_Death, Value) %>%
+#   rename(Number_of_Death = Value) %>%
+#   distinct(Year, Leading_Cause_Death, .keep_all = TRUE)
+# view(number_data)
+# 
+# # Subset for Ranking
+# ranking_data <- filter(raw_data, Characteristics == "Rank of leading causes of death") %>%
+#   select(Year, Leading_Cause_Death, Value) %>%
+#   rename(Ranking = Value)
+# 
+# 
+# # Identify causes of death with missing number of deaths or ranking
+# missing_number <- setdiff(unique(raw_data$Leading_Cause_Death), unique(number_data$Leading_Cause_Death))
+# missing_ranking <- setdiff(unique(raw_data$Leading_Cause_Death), unique(ranking_data$Leading_Cause_Death))
+# 
+# # Combine missing information
+# missing_info <- union(missing_number, missing_ranking)
+# 
+# print(missing_info)
+# 
+# # Create a full list of unique combinations of Year and Leading_Cause_Death
+# all_combinations <- expand.grid(Year = unique(raw_data$Year), Leading_Cause_Death = unique(raw_data$Leading_Cause_Death))
 
-# Subset for Number of Deaths
-rd_1 <- subset(raw_data, Characteristics == "Number of deaths") 
-rd_1$Number_Death <- rd_1$Value 
-rd_1 <- rd_1 %>% 
-  select(Year, Leading_Cause_Death, Number_Death)
+# Prepare datasets for "Number of deaths" and "Rank of leading causes of death"
+number_of_deaths <- raw_data %>%
+  filter(Characteristics == "Number of deaths") %>%
+  select(Year, Leading_Cause_Death, Number_of_Death = Value)
 
[…]

### 2024-03-14 scripts/02-data_cleaning.R
Update data cleaning script

@@ -18,60 +18,26 @@ raw_data <- read.csv("inputs/data/raw_data.csv")
 col_name <- names(raw_data)
 raw_data <- raw_data %>%
   rename("Year" = "REF_DATE",
-         "Leading_Cause_Death" = "Leading.causes.of.death..ICD.10.",
+         "Cause" = "Leading.causes.of.death..ICD.10.",
          "Value" = "VALUE") %>%
-  select("Year", "Characteristics", "Leading_Cause_Death", "Value")
-# 
-# # Subset for Number of Deaths
-# number_data <- filter(raw_data, Characteristics == "Number of deaths") %>%
-#   select(Year, Leading_Cause_Death, Value) %>%
-#   rename(Number_of_Death = Value) %>%
-#   distinct(Year, Leading_Cause_Death, .keep_all = TRUE)
-# view(number_data)
-# 
-# # Subset for Ranking
-# ranking_data <- filter(raw_data, Characteristics == "Rank of leading causes of death") %>%
-#   select(Year, Leading_Cause_Death, Value) %>%
-#   rename(Ranking = Value)
-# 
-# 
-# # Identify causes of death with missing number of deaths or ranking
-# missing_number <- setdiff(unique(raw_data$Leading_Cause_Death), unique(number_data$Leading_Cause_Death))
-# missing_ranking <- setdiff(unique(raw_data$Leading_Cause_Death), unique(ranking_data$Leading_Cause_Death))
-# 
-# # Combine missing information
-# missing_info <- union(missing_number, missing_ranking)
-# 
-# print(missing_info)
-# 
-# # Create a full list of unique combinations of Year and Leading_Cause_Death
-# all_combinations <- expand.grid(Year = unique(raw_data$Year), Leading_Cause_Death = unique(raw_data$Leading_Cause_Death))
-
-# Prepare datasets for "Number of deaths" and "Rank of leading causes of death"
+  select("Year", "Characteristics", "Cause", "Value") %>%
+  mutate(Cause = sub("\\[.*", "", Cause))
+  
 number_of_deaths <- raw_data %>%
   filter(Characteristics == "Number of deaths") %>%
-  select(Year, Leading_Cause_Death, Number_of_Death = Value)
+  select(Year, Cause, Death = Value)
 
 ranking_of_death <- raw_data %>%
   filter(Characteristics == "Rank of leading causes of death") %>%
-  select(Year, Leading_Cause_Death, Ranking = Value)
[…]

### 2024-03-15 scripts/02-data_cleaning.R
Updated data cleaning

@@ -20,8 +20,7 @@ raw_data <- raw_data %>%
   rename("Year" = "REF_DATE",
          "Cause" = "Leading.causes.of.death..ICD.10.",
          "Value" = "VALUE") %>%
-  select("Year", "Characteristics", "Cause", "Value") %>%
-  mutate(Cause = sub("\\[.*", "", Cause))
+  select("Year", "Characteristics", "Cause", "Value")
   
 number_of_deaths <- raw_data %>%
   filter(Characteristics == "Number of deaths") %>%
@@ -33,11 +32,10 @@ ranking_of_death <- raw_data %>%
 
 # Combine the subsets 
 combined_data <- full_join(number_of_deaths, ranking_of_death, by = c("Year", "Cause"))
-view(combined_data)
-
 
 # Filter out rows containing yearly totals
-cleaned_data <- filter(combined_data, !grepl("\\[A00-Y89\\]", Cause))
+cleaned_data <- filter(combined_data, !grepl("\\[A00-Y89\\]", Cause)) %>%
+  mutate(Cause = sub("\\[.*", "", Cause))
 view(cleaned_data)
 
 #### Save data ####

### 2024-03-15 outputs/paper/paper.qmd
Updated the models

@@
 library(knitr)
+library(rstanarm)
+library(modelsummary)
 
@@
 ```{r}
-#| label: Table 1
+#| label: Tbl-1
 #| fig-cap: Top-ten causes of death in Canada in 2022
@@
 
-combined_data %>%
-  filter(Year == 2021, Ranking <= 10) %>%
-  mutate(Number_of_Death = format(Number_of_Death, big.mark = ",")) %>%
+cleaned_data <- read.csv("../data/cleaned_data.csv")
+
+transformed_data <- cleaned_data %>%
+  add_count(Cause) %>%
+  filter(Year == 2022, Ranking <= 10) %>%
+  arrange(Ranking) %>%
+  mutate(Death = format(Death, big.mark = ","))
+
+transformed_data %>%
   kable(
-    col.names = c("Year", "Leading_Cause_Death", "Ranking", "Number_of_Death"),
-    align = c("l", "r", "r", "r"),
+    col.names = c("Year", "Cause", "Death", "Ranking", "Counts"),
+    align = c("l", "r", "r", "r", "r"),
     digits = 0, booktabs = TRUE, linesep = ""
   )
+
 ```
@@
 Talk more about it.
+
+```{r}
+#| label: Tbl-2
+#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
+#| echo: false
+
+```
 
@@
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: Fig-1
+#| fig-cap: Annual number of deaths for the top-five causes in 2021, since 2001, for Canada
 #| echo: false
@@
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+canada_top_five <-
+  transformed_data |>
+  filter(
+    Year == 2022, 
+    n == 22
+  ) |>
+  slice_max(order_by = desc(Ranking), n = 5) |>
+  pull(Cause)
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
+filtered_data <- cleaned_data |>
+  filter(Cause %in% canada_top_five)
+
+filtered_data |>
+  ggplot(aes(x = Year, y = Death, color = Cause)) +
+  geom_line() +
   theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+  labs(x = "Year", y = "Annual number of deaths in Canada") +
+  facet_wrap(vars(Cause), dir = "v", ncol = 1) +
+  theme(legend.position = "none")
+
 ```
@@
 Talk way more about it. 
+```{r}
+#| label: tbl-2
+#| fig-cap: Summary statistics of the number of yearly deaths, by cause, in Alberta, Canada
+#| echo: false
+#| warning: false
+#| message: false
 
+```
+
+```{r}
+#| label: tbl-3
+#| fig-cap: Modeling the most prevalent cause of deaths in Canada, 2001-2022
+#| echo: false
+#| warning: false
+#| message: false
+
+canada_death_poisson <- readRDS("../models/first_model.rds")
+canada_death_negbinomial <- readRDS("../models/first_model.rds")
+
+coef_short_names <- 
+  c("(Intercept)" = "(Intercept)",
+    "CauseCerebrovascular diseases " = "Cerebrovascular diseases",
+    "CauseChronic lower respiratory diseases " = "Chronic lower respiratory diseases",
+    "CauseDiseases of heart " = "Diseases of heart",
+    "CauseMalignant neoplasms " = "Malignant neoplasms"
+  )
+
+modelsummary(
+  list(
+    "Poisson" = canada_death_poisson,
+    "Negative binomial" = canada_death_negbinomial
+  ),
+  coef_map = coef_short_names,
+  statistic = "std.error"
+)
+```
 # Model
 
-The goal of our modelling strategy is twofold. Firstly,...
+The goal of our modelling strategy is twofold. 
+Firstly,...
+we would like to model the mortality in Canada.
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Here we briefly describe the Bayesian analysis model used to investigate...
+
+We model the incidences of death by one of the leading cause of deaths using a Poisson regression model and a negative Binomial regression model.
+Background details and diagnostics are included in [Appendix -@sec-model-details].
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+In our Poisson model, define $y_i$ as the number of deaths in a year due to a leading cause of death.
+Then $\lambda_i$ is the average rate of deaths due to this cause per year.
+$\lambda_i$ is linked to the predictor $x_i$ for the $i$th observation by a log link function where $\beta_0$ is the intercept and $\beta_1$ is the coefficient.
+We specifiy prior distributions for the parameters $\beta_0$ and $\beta_1$ in Bayesian analysis.
+We choose $\beta_0$ and $\beta_1$ to follow a normal distribution with mean 0 and conservative standard deviation of 2.5.
 
 \begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
+y_i&\sim \text{Poisson}(\lambda_i) \quad \text{(1)} \\
+\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
+\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
+\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
 \end{align}
@@
 
+In our negative binomial model, similarly define $y_i$ as the number of deaths due to a leading cause of death in a year.
+Then $\mu_i$ is the average count of deaths in a year.
+$\mu_i$ is linked to the predictor cause of death $x_i$ by a log link function with the intercept $\beta_0$ and the coefficient $\beta_i$.
+$\phi_i$ is the dispersion of the distribution measuring the extent of deviation from the count expected under a Poisson distribution.
+We specifiy the prior distributions for $\beta_0$ and $\beta_1$ to follow a normal distribution with mean 0 and conservative standard deviation of 2.5.
+We specify the prior of $\phi_i$ to follow a gamma distribution.
+
+\begin{align} 
+y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad \text{(1)} \\
+\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
+\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
+\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
+\phi &\sim \text{Gamma}(2, 0.1) \quad \text{5} \\
+\end{align}
+
 ### Model justification
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
+Since the residual of the number of deaths by a leading cause in a year does not follow a normal distribution, we consider using a generalized linear regression model.
+Since mortality occurrences are discrete events, we consider the Poisson regression model and its variant, negative binomial regression model, both of which relate explanatory variables to dependent variables representing counts of events (Reference: Poisson for aggregate data).
+Table 2 shows that the mean and variance are not the same.
+This indicates dispersion in the data, therefore negative binomial regression model may be more suitable.
 

### 2024-03-15 scripts/04-model.R

@@ -11,25 +11,43 @@ library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("outputs/data/cleaned_data.csv")
+analysis_data <- cleaned_data |>
+  filter(Cause %in% canada_top_five)
 
 ### Model data ####
-first_model <-
+
+#Poisson Model
+canada_death_poisson <-
   stan_glm(
-    formula = flying_time ~ length + width,
+    Death ~ Cause,
     data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    family = poisson(link = "log"),
     seed = 853
   )
 
+#N binomial Model
+canada_death_negbinomial <-
+  stan_glm(
+    Death ~ Cause,
+    data = analysis_data,
+    family = neg_binomial_2(link = "log"),
+    seed = 853
+  )
 
 #### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
-)
+
+saveRDS(canada_death_poisson, "outputs/models/canada_death_poisson.rds")
+saveRDS(canada_death_negbinomial, "outputs/models/canada_death_negbinomial.rds")
+
+
+# For the Poisson model
+poisson_terms <- names(coef(canada_death_poisson))
+
+# For the Negative Binomial model
+negbin_terms <- names(coef(canada_death_negbinomial))
+
+# Printing out the terms
+print(poisson_terms)
+print(negbin_terms)
 
 

### 2024-03-16 outputs/paper/paper.qmd
Updated model set up

@@
 
-The remainder of this paper is structured as follows. @sec-data....
+The remainder of this paper is structured as follows. In @sec-data, we showed the data and methodology utilized in this study. @sec-model discusses the model used to analyze the data
 
 # Data {#sec-data}
-
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
 
@@
 transformed_data <- cleaned_data %>%
+  filter(Ranking <= 10) %>%
   add_count(Cause) %>%
-  filter(Year == 2022, Ranking <= 10) %>%
+  filter(Year == 2021) %>%
   arrange(Ranking) %>%
@@
 
-filtered_data <- cleaned_data |>
+analysis_data <- cleaned_data |>
   filter(Cause %in% canada_top_five)
 
-filtered_data |>
+analysis_data |>
   ggplot(aes(x = Year, y = Death, color = Cause)) +
@@
 
-canada_death_poisson <- readRDS("../models/first_model.rds")
-canada_death_negbinomial <- readRDS("../models/first_model.rds")
+canada_death_poisson <- readRDS("../models/canada_death_poisson.rds")
+canada_death_negbinomial <- readRDS("../models/canada_death_negbinomial.rds")
 
-coef_short_names <- 
-  c("(Intercept)" = "(Intercept)",
+coef_short_names <- c(
+  "(Intercept)" = "(Intercept)",
   "CauseCerebrovascular diseases " = "Cerebrovascular diseases",
@@
 )
+
 ```
@@
 ```
+Table X presents the coefficients for various causes from both the Poisson and the Negative Binomial regression models.
+The intercept represents the baseline of the model, in this case, Accidents (unintentional injuries).
+It shows the log count of deaths when all other causes of death are at this reference level, i.e.
+when no one other specific cause is considered.
+
+The coefficients for specific causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of heart, and malignant neoplasms reflect the log-relative difference in the count of deaths relative to the reference level.
+Since we used a log link function, the coefficients represent effects on the log scale.
+Positive coefficients therefore indicate positive association between the cause and the number of deaths.
+
+Table shows that among the top five leading causes of death in Canada, diseases of the heart with a coefficient of 1.454 and malignant neoplasms with a coefficient of approximately 1.805 significantly indicates a strong positive association in both models.
+The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model.
+Exponentiating the coefficient to obtain exp(1.454) = 4.3 suggests that the expected count of deaths due to Diseases of the heart is about 3 times higher than the baseline category Accidents (unintentional injuries).
+Similarly exponentiating the coefficient for malignant neoplasms to obtain exp(1.805) shows the expected count of deaths is more than 5 times higher Accidents (unintentional injuries).
+Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about 7% less.
+Both the Poisson model and the negative binomial model yielded similar coefficients.
+
+This similarity suggests that both models are capturing similar relationship between the causes and the number of deaths.
+Comparing the Log Likelihood, The Negative Binomial model has a much higher (less negative) value (-1050.577) than the Poisson model (-16957.794).
+A higher log likelihood indicates a better model fit thus our analysis suggests the Negative Binomial model fits the data better.
+The other measures ELPD, LOOIC WAIC and RMSE all suggest that the negative binomial model fits the data better.
+(Appendix B)
 

### 2024-03-16 outputs/paper/paper.qmd
Update the introduction, data and appendix section of the paper

@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
-
 The remainder of this paper is structured as follows. In @sec-data, we showed the data and methodology utilized in this study. @sec-model discusses the model used to analyze the data
@@
 # Data {#sec-data}
+
+R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data cleaning and visualization.
 
@@
 ```
-
-Talk more about it.
-
-```{r}
-#| label: Tbl-2
-#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
-#| echo: false
-
-```
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
 
@@
 
-Talk way more about it. 
 ```{r}
-#| label: tbl-2
-#| fig-cap: Summary statistics of the number of yearly deaths, by cause, in Alberta, Canada
+#| label: Tbl-2
+#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
 #| echo: false
-#| warning: false
-#| message: false
 
+datasummary(Death ~ Min + Mean + Max + SD + Var + N,
+            fmt = 0,
+            data = analysis_data)
 ```
+@tbl-2 shows that the mean, 32,603, is different to the variance 673,243,267.
 
@@
 Background details and diagnostics are included in [Appendix -@sec-model-details].
+
+R packages rstanarm[@rstanarm] and modelsummary [@model] are used to build and analyze the models.
 
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
 Table X presents the coefficients for various causes from both the Poisson and the Negative Binomial regression models.
@@
 The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model.
-Exponentiating the coefficient to obtain exp(1.454) = 4.3 suggests that the expected count of deaths due to Diseases of the heart is about 3 times higher than the baseline category Accidents (unintentional injuries).
-Similarly exponentiating the coefficient for malignant neoplasms to obtain exp(1.805) shows the expected count of deaths is more than 5 times higher Accidents (unintentional injuries).
-Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about 7% less.
+Exponentiating the coefficient 1.454 to obtain `r round(exp(1.454),2)` suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries).
+Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is `r round(exp(1.805)-1, 2)` times higher Accidents (unintentional injuries).
+Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about `r round(1-exp(-0.070),2)` less.
 Both the Poisson model and the negative binomial model yielded similar coefficients.
@@
 This similarity suggests that both models are capturing similar relationship between the causes and the number of deaths.
-Comparing the Log Likelihood, The Negative Binomial model has a much higher (less negative) value (-1050.577) than the Poisson model (-16957.794).
+Comparing the Log Likelihood, The Negative Binomial model has a much higher (less negative) value -1050.577 than the Poisson model -16957.794.
 A higher log likelihood indicates a better model fit thus our analysis suggests the Negative Binomial model fits the data better.
-The other measures ELPD, LOOIC WAIC and RMSE all suggest that the negative binomial model fits the data better.
-(Appendix B)
+The other measures ELPD, LOOIC WAIC and RMSE all suggest that the negative binomial model fits the data better @sec-model-details
 
+```{r}
+
+poisson <- kfold(canada_death_poisson, k = 10)
+neg_binomial <- kfold(canada_death_negbinomial, k = 10)
+
+loo_compare(poisson, neg_binomial)
+
+```
 # Discussion
@@
 
-# Additional data details
+# Additional data details {#sec-data-details}
+
+```{r}
+#| label: tbl-5
+#| fig-cap: A sample of the cleaned dataset, 2001-2022
+#| echo: false
+#| warning: false
+#| message: false
+sampled_data <- sample_n(cleaned_data, 5) %>%
+  kable()
+sampled_data
+```
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check for the Poisson regression model described in @sec-4.
+This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align reasonably well with the actual data (represented by the dark blue line) with minor misalignment.
+This suggests the model represent the data well.
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the negative binomial regression model described in @sec-4.
+This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align generally well with the actual data (represented by the dark blue line).
+This suggests the model represents the data well.
 
@@
 #| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models"
+#| fig-subcap: ["Poisson model", "Negative binomial model"]
 
-pp_check(first_model) +
-  theme_classic() +
+pp_check(canada_death_poisson) +
   theme(legend.position = "bottom")
 
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
+pp_check(canada_death_negbinomial) +
+  theme(legend.position = "bottom")
 ```
@@
 ## Diagnostics
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheckandposteriorvsprior
+#| layout-ncol: 2
+#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models"
+#| fig-subcap: ["Poisson model", "Negative binomial model"]
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+pp_check(canada_death_poisson) +
+  theme(legend.position = "bottom")
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+pp_check(canada_death_negbinomial) +
+  theme(legend.position = "bottom")
+```
+
+@fig-stanareyouokay-1 is a trace plot for the Poisson regression model.
+It shows a horizontal, dense band of samples without any systematic patterns, drifts, or long periods of stagnation.
+This pattern suggests that the chain is mixing well and sampling efficiently from the posterior distribution.
+This suggests the Poisson regression model is suitable.
+
+@fig-stanareyouokay-2 is a Rhat plot. It shows an Rhat value close to 1.  This suggests that the chains have converged to the target distribution.
 
@@
 #| warning: false
-#| label: fig-stanareyouokay
+#| label: fig-stanareyouokay-a
 #| fig-cap: "Checking the convergence of the MCMC algorithm"
@@
 
-plot(first_model, "trace")
+plot(canada_death_poisson, "trace")
+plot(canada_death_poisson, "rhat")
 
-plot(first_model, "rhat")
 ```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-stanareyouokay-b
+#| fig-cap: "Checking the convergence of the MCMC algorithm"
+#| fig-subcap: ["Trace plot", "Rhat plot"]
+#| layout-ncol: 2
+
+plot(canada_death_negbinomial, "trace")
+plot(canada_death_negbinomial, "rhat")
+```
+
+@fig-stanareyouokay-b-1 is a trace plot for the negative binomial regression model.
+It also shows a horizontal, dense band of samples without abnormalities suggesting the chain is mixing well and sampling efficiently from the posterior distribution.
+This suggests the negative binomial model is suitable.
+
+@fig-stanareyouokay-b-2 is a Rhat plot for the negative binomial regression model.
+It shows an Rhat value close to 1.
+This again suggests that the chains have converged to the target distribution in the negative regression model.
 

### 2024-03-16 scripts/02-data_cleaning.R

@@ -13,6 +13,7 @@ library(dplyr)
 
 ### Clean the data ###
 raw_data <- read.csv("inputs/data/raw_data.csv")
+dim(raw_data)
 
 # # Select the columns of interest
 col_name <- names(raw_data)

### 2024-03-16 outputs/paper/paper.qmd
Updated references

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Decoding Mortality: Trends and Causes in Canada between 2001-2022"
+subtitle: "A Bayesian Statistical Analysis Unveiling the Predominance of Malignant Neoplasms and Heart Diseases"
 author: 
-  - First author
-  - Another author
-thanks: "Code and data are available at: LINK."
+  - Yuchao Niu
+thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
 date: today
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "The study of mortality offers key public health insights into the fundamental causes and trends of death.
+This research investigates Canada's mortality trends from 2001 to 2022, employing Bayesian Poisson and negative binomial regression models to analyze data from the Canadian Vital Statistics Death Database.
+It reveals that malignant neoplasms and heart diseases significantly elevate mortality rates, surpassing the effects of unintentional accidents by five times and three times, respectively.
+Such findings highlight the critical necessity for targeted public health initiatives focusing on the prevention, early detection, and treatment of these conditions."
 format: pdf

### 2024-03-16 outputs/paper/paper.qmd
Update paper

@@
 library(modelsummary)
+library(arrow)
 
@@
 
-The remainder of this paper is structured as follows. In @sec-data, we showed the data and methodology utilized in this study. @sec-model discusses the model used to analyze the data
+Death is a dreaded yet unavoidable event in life.
+As Benjamin Franklin famously stated, "In this world, nothing can be said to be certain, except death and taxes." Population mortality has been meticulously recorded over the past century, enabling the study of epidemiological trends, such as those observed in the recent pandemic.
+Crucially, analyzing region-specific mortality and identifying the leading causes of death inform public health decisions and guide the effective allocation of government resources to prevent and treat diseases.
+The availability of vast data sources unveils patterns and valuable information regarding causes of death.
+
+This study investigates mortality in Canada from 2001 to 2022, utilizing the Canadian Vital Statistics Death Database by Statistics Canada.
+The focus is on identifying the risk factors associated with each cause of death.
+We employed a Bayesian approach, using both the Poisson regression model and the Negative Binomial regression model, to analyze the top five leading causes of mortality in Canada.
+The Bayesian model applies Bayesian inference to continually update the probability of a hypothesis as more data becomes available.
+The Poisson regression model is traditionally favored for count data but assumes equal mean and variance, an assumption that does not always hold in mortality data due to over-dispersion.
+The Negative Binomial regression model addresses this by accounting for over-dispersion, offering a more flexible approach for such data.
+
+The top five leading causes of death identified are malignant neoplasm, diseases of the heart, chronic respiratory diseases, cerebrovascular disease, and unintentional accidents.
+Malignant neoplasm exhibits the strongest impact death counts, being 5 times more influential than accidents.
+Diseases of the heart account for 3 times more deaths than unintentional accidents.
+The findings advocate for prioritizing public health resources towards the screening, prevention, and treatment of these conditions.
+
+The remainder of this paper is organized as follows.
+@sec-data discusses the data and measurement methods used in this study.
+@sec-model presents the model utilized in the analysis.
+@sec-result presents the results, and @sec-discussion discusses the implications, limitations, and proposes future directions for research.
 
@@
 
-R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data cleaning and visualization.
+## Data Source
+
+This study was inspired by the analysis of mortality in Alberta [@story], focusing on the leading causes of death, and extends the analysis to Canada, exploring the leading causes of death nationwide.
+The dataset for this analysis is sourced from the open government portal of Statistics Canada, summarizing the causes of death and the number of individuals affected across all age groups from 2000 to 2022.
+This data originates from the Canadian Vital Statistics – Death Database (CVSD), an administrative survey that collects demographic and medical information, including the cause of death, from vital statistics registries in all provinces and territories, a practice ongoing since 1921.
+
+## Data Measurement
+
+The target population of the Canadian Vital Statistics – Death database (CVSD) includes deaths occurring in Canada among both Canadian residents and non-residents.
+As the survey operates on a census with a cross-sectional design, it does not employ sampling methods.
+
+Data on registered deaths is submitted to Statistics Canada by each provincial and territorial Vital Statistics Registry.
+The death registration form comprises personal details provided to the funeral director by someone knowledgeable about the deceased and a medical certificate of cause of death, completed by the last attending medical professional or a coroner in cases necessitating an inquest or inquiry.
+
+The data collected includes age, sex, marital status, residence and birthplace of the deceased, date of death, underlying cause of death according to the World Health Organization's International Statistical Classification of Diseases and Related Health Problems (ICD), province or territory of death occurrence, place of accident for non-transport accidental deaths, and autopsy details.
+
+Statistics Canada conducts routine quality checks to ensure an error rate below 3%.
+However, potential biases may arise from the classification of death causes, adherence to the ICD, inclusion of resident data, and data transmission methods.
+Specifically, the shift in data collection for Canadian residents dying in the United States post-2009, the unavailability of Yukon data post-2017, and the transition to electronic data transmission via the National Routing System (NRS) could introduce biases.
+These factors may affect the survey's representativeness and accuracy, which are crucial for researchers to consider.
+
+## Data Characteristics
+
+The original dataset obtained from the Statistics Canada open government portal includes 2,268 observations across 18 variables, offering detailed demographic and geographical information on registered deaths.
+I aim to focus on five specific variables: 'Reference Period', 'Leading Causes of Death (ICD-10)', 'Characteristics', 'Ranking', and 'Number of Deaths'.
+The 'Reference Period' denotes the year of death registration (2000-2022), while 'Leading Causes of Death' categorizes the cause according to the International Statistical Classification of Diseases and Related Health Problems (ICD-10).
+'Characteristics' encompasses both the 'Number of Deaths' and their 'Ranking', indicating either the total deaths or the disease's rank in a given year, respectively.
+A sample of the cleaned data can be found in @sec-data-details.
+
+## Data Visualization
+
+@tbl-1 highlights the top ten causes of death in 2022, with counts reflecting their frequency over 22 years.
+Except for COVID-19, all other causes have consistently been among the top ten annually.
+We selected the top five for detailed examination.
+
+@fig-1 depicts the annual death trends for these top five causes.
+There has been a modest increase in deaths from Accidents (unintentional injuries) over two decades.
+Deaths due to Malignant Neoplasms have risen from 60,000 in the early 2000s to nearly 80,000 in 2023, while Cerebrovascular Diseases, Chronic Lower Respiratory Diseases, and Diseases of the Heart have maintained steady annual deaths at approximately 25,000, 20,000, and 60,000, respectively.
+
+R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], arrow[@arrow], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data downloading, cleaning and visualization.
 
 ```{r}
-#| label: Tbl-1
-#| fig-cap: Top-ten causes of death in Canada in 2022
-#| echo: false
 
@@
 ```
-
 ```{r}
-#| label: Fig-1
+#| label: fig-1
 #| fig-cap: Annual number of deaths for the top-five causes in 2021, since 2001, for Canada
@@
 
-canada_top_five <-
+top_five <-
   transformed_data |>
@@
 
-analysis_data <- cleaned_data |>
-  filter(Cause %in% canada_top_five)
+# analysis_data <- cleaned_data |>
+#   filter(Cause %in% canada_top_five)
 
-analysis_data |>
+cleaned_data |>
+  filter(Cause %in% top_five)|>
   ggplot(aes(x = Year, y = Death, color = Cause)) +
@@
 ```{r}
-#| label: Tbl-2
+#| label: fig-2
+#| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
+#| warning: false
+#| message: false
+
+cleaned_data |>
+  filter(Year == 2021)
+  ggplot(top_five, aes(x="", y=total_deaths, fill=cause_of_death)) +
+  geom_bar(width = 1, stat = "identity") +
+  coord_polar("y") +
+  theme_void() +
+  labs(fill="Cause of Death")
+
+```
+
+# Model {#sec-model}
+
+The goal of our modeling strategy is to analyze mortality data for the leading causes of death in Canada over the past two decades, aiming to understand the relationship between disease and the number of deaths.
+We employ a Bayesian approach, assigning prior distributions to parameters based on prior knowledge and specifying a likelihood function based on the observed data.
+By applying Bayes' theorem, we update our prior beliefs to form posterior distributions of the parameters.
+Markov Chain Monte Carlo (MCMC) methods are utilized to approximate these posterior distributions, allowing us to integrate prior knowledge and interpret our findings within a probabilistic framework.
+
+## Model set-up
+
+In our Poisson model, define $y_i$ as the number of deaths in a year due to a leading cause of death.
+Then $\lambda_i$ is the average rate of deaths due to this cause per year.
+$\lambda_i$ is linked to the predictor $x_i$ for the $i$th observation by a log link function where $\beta_0$ is the intercept and $\beta_1$ is the coefficient.
+We specifiy prior distributions for the parameters $\beta_0$ and $\beta_1$ in Bayesian analysis.
+We choose $\beta_0$ and $\beta_1$ to follow a normal distribution with mean 0 and conservative standard deviation of 2.5.
+
+\begin{align} 
+y_i&\sim \text{Poisson}(\lambda_i) \quad \text{(1)} \\
+\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
+\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
+\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
+\end{align}
+
+In our negative binomial model, similarly define $y_i$ as the number of deaths due to a leading cause of death in a year.
+Then $\mu_i$ is the average count of deaths in a year.
+$\mu_i$ is linked to the predictor cause of death $x_i$ by a log link function with the intercept $\beta_0$ and the coefficient $\beta_i$.
+$\phi_i$ is the dispersion of the distribution measuring the extent of deviation from the count expected under a Poisson distribution.
+We specifiy the prior distributions for $\beta_0$ and $\beta_1$ to follow a normal distribution with mean 0 and conservative standard deviation of 2.5.
+We specify the prior of $\phi_i$ to follow a gamma distribution.
+
+\begin{align} 
+y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad \text{(1)} \\
+\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
+\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
+\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
+\phi &\sim \text{Gamma}(2, 0.1) \quad \text{5} \\
+\end{align}
+
+### Model justification
+
+Since the residual of the number of deaths by a leading cause in a year does not follow a normal distribution, we consider using a generalized linear regression model.
+Since mortality occurrences are discrete events, we consider the Poisson regression model and its variant, negative binomial regression model, both of which relate explanatory variables to dependent variables representing counts of events.
+@tbl-2 shows that the mean, 32,603, is different from the variance, 673,243,267.
+This indicates dispersion in the data, therefore negative binomial regression model may be more suitable.
+
+```{r}
+#| label: tbl-2
 #| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
@@
 ```
-@tbl-2 shows that the mean, 32,603, is different to the variance 673,243,267.
+
+For model justification, we acknowledge that mortality occurrences are discrete events, making the Poisson and negative binomial regression models suitable for relating explanatory variables to event counts.
+The negative binomial model is particularly chosen for its capacity to handle overdispersion, a common feature in mortality data, under the assumption that the model's residuals are uncorrelated.
+
+We model the incidences of death by one of the leading cause of deaths using a Poisson regression model and a negative Binomial regression model.
+Background details and diagnostics are included in [@sec-model-details].
+
+R packages rstanarm[@rstanarm] and modelsummary [@model] are used to build and analyze the models. We use the default priors from `rstanarm`.
+
+# Results {#sec-result}
+
+@tbl-3 presents coefficients from both Poisson and Negative Binomial regression models for various causes of death.
+The intercept, representing the baseline of Accidents (unintentional injuries), indicates the log count of deaths when no other specific cause is considered.
+Coefficients for causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of the heart, and malignant neoplasms reflect log-relative differences in death counts relative to Accidents.
+Positive coefficients indicate a positive association with the number of deaths.
 
@@
 ```
-# Model
 
-The goal of our modelling strategy is twofold. 
-Firstly,...
-we would like to model the mortality in Canada.
-
-Here we briefly describe the Bayesian analysis model used to investigate...
-
-We model the incidences of death by one of the leading cause of deaths using a Poisson regression model and a negative Binomial regression model.
-Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-R packages rstanarm[@rstanarm] and modelsummary [@model] are used to build and analyze the models.
-
-## Model set-up
-
-In our Poisson model, define $y_i$ as the number of deaths in a year due to a leading cause of death.
-Then $\lambda_i$ is the average rate of deaths due to this cause per year.
-$\lambda_i$ is linked to the predictor $x_i$ for the $i$th observation by a log link function where $\beta_0$ is the intercept and $\beta_1$ is the coefficient.
-We specifiy prior distributions for the parameters $\beta_0$ and $\beta_1$ in Bayesian analysis.
-We choose $\beta_0$ and $\beta_1$ to follow a normal distribution with mean 0 and conservative standard deviation of 2.5.
-
-\begin{align} 
-y_i&\sim \text{Poisson}(\lambda_i) \quad \text{(1)} \\
-\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
-\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
-\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-In our negative binomial model, similarly define $y_i$ as the number of deaths due to a leading cause of death in a year.
-Then $\mu_i$ is the average count of deaths in a year.
-$\mu_i$ is linked to the predictor cause of death $x_i$ by a log link function with the intercept $\beta_0$ and the coefficient $\beta_i$.
-$\phi_i$ is the dispersion of the distribution measuring the extent of deviation from the count expected under a Poisson distribution.
-We specifiy the prior distributions for $\beta_0$ and $\beta_1$ to follow a normal distribution with mean 0 and conservative standard deviation of 2.5.
-We specify the prior of $\phi_i$ to follow a gamma distribution.
-
-\begin{align} 
-y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad \text{(1)} \\
-\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
-\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
-\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
-\phi &\sim \text{Gamma}(2, 0.1) \quad \text{5} \\
-\end{align}
-
-### Model justification
-
-Since the residual of the number of deaths by a leading cause in a year does not follow a normal distribution, we consider using a generalized linear regression model.
-Since mortality occurrences are discrete events, we consider the Poisson regression model and its variant, negative binomial regression model, both of which relate explanatory variables to dependent variables representing counts of events (Reference: Poisson for aggregate data).
-Table 2 shows that the mean and variance are not the same.
-This indicates dispersion in the data, therefore negative binomial regression model may be more suitable.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-Table X presents the coefficients for various causes from both the Poisson and the Negative Binomial regression models.
-The intercept represents the baseline of the model, in this case, Accidents (unintentional injuries).
-It shows the log count of deaths when all other causes of death are at this reference level, i.e.
-when no one other specific cause is considered.
-
-The coefficients for specific causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of heart, and malignant neoplasms reflect the log-relative difference in the count of deaths relative to the reference level.
-Since we used a log link function, the coefficients represent effects on the log scale.
-Positive coefficients therefore indicate positive association between the cause and the number of deaths.
+For instance, @tbl-3 shows that diseases of the heart have a coefficient of 1.454 in both models, suggesting the expected death count from this cause is approximately `r round(exp(1.454),2)`  times higher than from Accidents (unintentional injuries), after exponentiating the coefficient.
+Similarly, malignant neoplasms have a coefficient of approximately 1.805, indicating an expected count more than `r round(exp(1.805)-1, 2)` times higher than Accidents.
+Conversely, chronic lower respiratory diseases show a modest decrease in mortality compared to Accidents, with a coefficient of -0.070 in the Poisson model, indicating about `r round(1-exp(-0.070),2)` fewer deaths.
 
@@
 The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model.
-Exponentiating the coefficient 1.454 to obtain `r round(exp(1.454),2)` suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries).
-Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is `r round(exp(1.805)-1, 2)` times higher Accidents (unintentional injuries).
-Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about `r round(1-exp(-0.070),2)` less.
-Both the Poisson model and the negative binomial model yielded similar coefficients.
+Exponentiating the coefficient 1.454 to obtain suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries).
+Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is ` times higher Accidents (unintentional injuries).
+Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about  less.
 
-This similarity suggests that both models are capturing similar relationship between the causes and the number of deaths.
-Comparing the Log Likelihood, The Negative Binomial model has a much higher (less negative) value -1050.577 than the Poisson model -16957.794.
-A higher log likelihood indicates a better model fit thus our analysis suggests the Negative Binomial model fits the data better.
-The other measures ELPD, LOOIC WAIC and RMSE all suggest that the negative binomial model fits the data better @sec-model-details
+Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts.
+The Negative Binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794.
+The negative binomial model's superior fit is further supported by a higher ELPD value compared to the Poisson model.
+Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.
 
@@
 ```
-# Discussion
 
-## First discussion point {#sec-first-point}
+# Discussion {#sec-discussion}
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+## Findings {#sec-first-point}
 
-## Second discussion point
+Our analysis over the past 20 years reveals that the expected count of deaths from malignant neoplasm, the leading cause of death, is approximately `r round(exp(1.805),2)` times higher than that from accidents (unintentional).
+Diseases of the heart, ranking second, account for `r round(exp(1.454) - 1,2)` times more deaths than accidents.
+This contrasts with global statistics from the WHO, highlighting the significance of allocating resources towards cancer prevention and treatment in Canada [@who].
+The analysis of Alberta's data shows similarities with national trends, with the notable exception of organic dementia replacing accidents due to its correlation with aging, despite Alberta having the lowest percentage of senior population at 15.2% by 2023 [@aging].
+This calls for further investigation into Alberta's health indicators and lifestyle factors.
 
-## Third discussion point
+## Ethical implications
+
+The study ensures confidentiality and respectful data handling, particularly when dealing with rare causes of death.
+It raises concerns about potential disparities in resource access and safety, urging further research into occupational and transportation inequities.
+Sampling biases were minimized, but differential data access could lead to underrepresentation, notably for Yukon post-2017, potentially affecting policy benefits [@statcan]
 
@@
 
-Weaknesses and next steps should also be included.
+While the Poisson and Negative Binomial regression models demonstrate good fit, they assume independence of death counts, which may not reflect reality due to clustering by geographic and demographic factors.
+Additionally, we only considered one variable the leading cause of death, without considering rarer causes or potential confounders like socioeconomic factors.
+This constrains the ability of the model to capture the nuances of mortality and its drivers.
+
+Future studies could aim to address these weaknesses by incorporating spatial and demographic variables to account for clustering effects.
+Future studies can broaden the scope to include more regions, socioeconomical variables, and their interactions.
+Longitudinal studies could offer insights into mortality trends and the long-term effects of environmental factors like climate change.
 

### 2024-03-17 outputs/paper/paper.qmd
Updated the paper and references.bib

@@
 thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
-date: today
+date: yesterday
 date-format: long
@@
 number-sections: true
+toc: true
 bibliography: references.bib
@@
 library(modelsummary)
+library(gt)
 library(arrow)
+library(dplyr)
+library(broom.mixed)
 
@@
 The 'Reference Period' denotes the year of death registration (2000-2022), while 'Leading Causes of Death' categorizes the cause according to the International Statistical Classification of Diseases and Related Health Problems (ICD-10).
-'Characteristics' encompasses both the 'Number of Deaths' and their 'Ranking', indicating either the total deaths or the disease's rank in a given year, respectively.
+'Characteristics' encompasses both the 'Number of Deaths' and their 'Ranking', indicating either the total deaths or the cause's rank in a given year, respectively.
 A sample of the cleaned data can be found in @sec-data-details.
+
+R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], arrow[@arrow], knitr[@Knitr] , ggplot2[@ggplot2], broom[@broom], dplyr[@dplyr] and  have been used in data downloading, cleaning and visualization.
 
@@
 
-@fig-1 depicts the annual death trends for these top five causes.
-There has been a modest increase in deaths from Accidents (unintentional injuries) over two decades.
-Deaths due to Malignant Neoplasms have risen from 60,000 in the early 2000s to nearly 80,000 in 2023, while Cerebrovascular Diseases, Chronic Lower Respiratory Diseases, and Diseases of the Heart have maintained steady annual deaths at approximately 25,000, 20,000, and 60,000, respectively.
+```{r}
+#| label: tbl-1
+#| tbl-cap: "Annual number of deaths for the top-five causes in 2021, since 2001, for Canada"
+#| echo: false
+#| warning: false
+#| message: false
 
-R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], arrow[@arrow], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data downloading, cleaning and visualization.
-
-```{r}
-
-cleaned_data <- read.csv("../data/cleaned_data.csv")
+cleaned_data <- read_parquet("../data/cleaned_data.parquet")
 
@@
 ```
+
+@fig-1 depicts the annual death trends for these top five causes.
+There has been a modest increase in deaths from Accidents (unintentional injuries) over two decades.
+Deaths due to Malignant Neoplasms have risen from 60,000 in the early 2000s to nearly 80,000 in 2023, while Cerebrovascular Diseases, Chronic Lower Respiratory Diseases, and Diseases of the Heart have maintained steady annual deaths at approximately 25,000, 20,000, and 60,000, respectively.
+
 ```{r}
@@
 
-top_five <-
-  transformed_data |>
-  filter(
-    Year == 2022, 
-    n == 22
-  ) |>
-  slice_max(order_by = desc(Ranking), n = 5) |>
+top_five <- transformed_data |>
+  filter(n == 22) |>
+  slice_max(order_by = desc(Ranking), n = 5)
+
+top_five_causes <- top_five |>
   pull(Cause)
 
-# analysis_data <- cleaned_data |>
-#   filter(Cause %in% canada_top_five)
+analysis_data <- cleaned_data |>
+  filter(Cause %in% top_five_causes)
 
-cleaned_data |>
-  filter(Cause %in% top_five)|>
+analysis_data |>
   ggplot(aes(x = Year, y = Death, color = Cause)) +
@@
 #| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
+#| echo: false
 #| warning: false
@@
 
-cleaned_data |>
-  filter(Year == 2021)
-  ggplot(top_five, aes(x="", y=total_deaths, fill=cause_of_death)) +
-  geom_bar(width = 1, stat = "identity") +
-  coord_polar("y") +
-  theme_void() +
-  labs(fill="Cause of Death")
+# cleaned_data |>
+#   filter(Year == 2021) |>
+#   mutate(Percentage = Death / sum(Death) * 100) |>
+#   ggplot(top_five, aes(x="", y= Death, fill= Cause)) +
+#   geom_bar(width = 1, stat = "identity") +
+#   coord_polar("y") +
+#   theme_void() +
+#   labs(fill="Cause of Death")
 
@@
 \begin{align} 
-y_i&\sim \text{Poisson}(\lambda_i) \quad \text{(1)} \\
-\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
-\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
-\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
+y_i&\sim \text{Poisson}(\lambda_i) \quad \\
+\text{log}(\lambda_i) &= \beta_0 + \beta_1 X_i \quad \\
+\beta_0 &\sim \text{Normal}(0, 2.5) \quad \\
+\beta_1 &\sim \text{Normal}(0, 2.5) \quad 
 \end{align}
@@
 \begin{align} 
-y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad \text{(1)} \\
-\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \text{(2)} \\
-\beta_0 &\sim \text{Normal}(0, 2.5) \quad \text{(3)} \\
-\beta_1 &\sim \text{Normal}(0, 2.5) \quad \text{(4)} \\
-\phi &\sim \text{Gamma}(2, 0.1) \quad \text{5} \\
+y_i&\sim \text{NegativeBinomial}(\mu, \phi) \quad  \\
+\text{log}(\mu_i) &= \beta_0 + \beta_1 X_i \quad \\
+\beta_0 &\sim \text{Normal}(0, 2.5) \quad \\
+\beta_1 &\sim \text{Normal}(0, 2.5) \quad \\
+\phi &\sim \text{Gamma}(2, 0.1) \quad 
 \end{align}
@@
 #| label: tbl-2
-#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
+#| tbl-cap: "Summary Statistics of the number of yearly deaths, by cause, in Canada"
 #| echo: false
@@
 
-@tbl-3 presents coefficients from both Poisson and Negative Binomial regression models for various causes of death.
+@tbl-3 presents coefficients from both Poisson and negative binomial regression models for various causes of death.
 The intercept, representing the baseline of Accidents (unintentional injuries), indicates the log count of deaths when no other specific cause is considered.
@@
 ```{r}
+#|message= false
+#|warning= false
+#|include= false
+
+canada_death_poisson <- readRDS("../models/canada_death_poisson.rds")
+canada_death_negbinomial <- readRDS("../models/canada_death_negbinomial.rds")
+
+tidy_poisson <- tidy(canada_death_poisson)
+tidy_negbinomial <- tidy(canada_death_negbinomial)
+
+coef_short_names_corrected <- c(
+  "(Intercept)" = "(Intercept)",
+  "CauseCerebrovascular diseases" = "Cerebrovascular diseases",
+  "CauseChronic lower respiratory diseases" = "Chronic lower respiratory diseases",
+  "CauseDiseases of heart" = "Diseases of heart",
+  "CauseMalignant neoplasms" = "Malignant neoplasms",
+  "CauseCOVID-19" = "COVID-19"  # Ensure this matches the term exactly as it appears in your model output
+)
+
+# Apply the corrected mapping to rename terms for the Poisson model
+tidy_poisson$term <- coef_short_names_corrected[tidy_poisson$term]
+# Handling any terms that did not match by retaining their original names
+tidy_poisson$term[is.na(tidy_poisson$term)] <- names(coef_short_names_corrected)[is.na(tidy_poisson$term)]
+
+# Apply the corrected mapping to rename terms for the Negative Binomial model
+tidy_negbinomial$term <- coef_short_names_corrected[tidy_negbinomial$term]
+# Handling any terms that did not match by retaining their original names
+tidy_negbinomial$term[is.na(tidy_negbinomial$term)] <- names(coef_short_names_corrected)[is.na(tidy_negbinomial$term)]
+
+# Combine the data for comparison
+comparison <- bind_rows(
+  mutate(tidy_poisson, model = "Poisson"),
+  mutate(tidy_negbinomial, model = "Negative Binomial")
+)
+
+# Assuming 'comparison' is your final data frame from previous steps
+comparison_table <- comparison %>%
+  gt() %>%
+  tab_header(
+    title = "Model Comparison: Poisson vs. Negative Binomial"
+  ) %>%
+  cols_label(
+    term = "Term",
+    estimate = "Estimate",
+    std.error = "Standard Error",
+    model = "Model"
+  ) %>%
+  fmt_number(
+    columns = c(estimate, std.error),
+    decimals = 3
+  ) %>%
+  tab_spanner(
+    label = "Statistics",
+    columns = c(estimate, std.error)
+  ) %>%
+  tab_style(
+    style = cell_text(align = 'center'),
+    locations = cells_body(columns = everything())
+  ) %>%
+  tab_options(
+    heading.title.font.size = 14,
+    heading.subtitle.font.size = 12,
+    table.font.size = 12
+  )
+```
+
+```{r}
 #| label: tbl-3
-#| fig-cap: Modeling the most prevalent cause of deaths in Canada, 2001-2022
+#| tbl-cap: "Modeling the most prevalent cause of deaths in Canada, 2001-2022"
 #| echo: false
@@
 
-canada_death_poisson <- readRDS("../models/canada_death_poisson.rds")
-canada_death_negbinomial <- readRDS("../models/canada_death_negbinomial.rds")
-
-coef_short_names <- c(
-  "(Intercept)" = "(Intercept)",
-  "CauseCerebrovascular diseases " = "Cerebrovascular diseases",
-  "CauseChronic lower respiratory diseases " = "Chronic lower respiratory diseases",
-  "CauseDiseases of heart " = "Diseases of heart",
-  "CauseMalignant neoplasms " = "Malignant neoplasms"
-)
-
-modelsummary(
-  list(
-    "Poisson" = canada_death_poisson,
-    "Negative binomial" = canada_death_negbinomial
-  ),
-  coef_map = coef_short_names,
-  statistic = "std.error"
-)
-
+comparison_table
 ```
@@
 Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts.
-The Negative Binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794.
-The negative binomial model's superior fit is further supported by a higher ELPD value compared to the Poisson model.
+The negative binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794.
+The negative binomial model's better fit is further supported by a higher ELPD value compared to the Poisson model as shown in @tbl-4.
 Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.
@@
 ```{r}
+#| label: tbl-4
+#| tbl-cap: "Cross validation comparision between Poisson model and negative binomial model "
+#| echo: false
+#| warning: false
+#| message: false
 
@@
 
-loo_compare(poisson, neg_binomial)
+(loo_compare(poisson, neg_binomial))
 
@@
 
-While the Poisson and Negative Binomial regression models demonstrate good fit, they assume independence of death counts, which may not reflect reality due to clustering by geographic and demographic factors.
+While the Poisson and negative binomial regression models demonstrate good fit, they assume independence of death counts, which may not reflect reality due to clustering by geographic and demographic factors.
 Additionally, we only considered one variable the leading cause of death, without considering rarer causes or potential confounders like socioeconomic factors.
@@
 
+The raw dataset was downloaded from Statistics Canada open data portal and cleaned by selecting the variables of interest "Year", "Causes of Death", "Characteristics" and "Value".
+The "Characteristics" variable is converted into a new variable "Death" indicating the number of death attributed to the cause and a new "Rank" variable showing the rank of the cause in a given year.
+@tbl-5 presents a sample of the cleaned dataset.
+
 ```{r}
 #| label: tbl-5
-#| fig-cap: A sample of the cleaned dataset, 2001-2022
+#| tbl-cap: "A sample of the cleaned dataset of number of death attributed to causes of death in Canada, 2001-2022"
 #| echo: false
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check for the Poisson regression model described in @sec-4.
+In @fig-ppcheck-a-1 we implement a posterior predictive check for the Poisson regression model described in @sec-result.
 This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align reasonably well with the actual data (represented by the dark blue line) with minor misalignment.
@@
 
-In @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the negative binomial regression model described in @sec-4.
-This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align generally well with the actual data (represented by the dark blue line).
+In @fig-ppcheck-a-2 we implement a posterior predictive check for the negative binomial regression model described in @sec-result This shows predictions generated from the posterior distribution of the model parameters (represented by the light blue lines) align generally well with the actual data (represented by the dark blue line).
 This suggests the model represents the data well.
@@
 #| warning: false
-#| label: fig-ppcheckandposteriorvsprior
+#| label: fig-ppcheck-a
 #| layout-ncol: 2
@@
 ## Diagnostics
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models"
-#| fig-subcap: ["Poisson model", "Negative binomial model"]
 
-pp_check(canada_death_poisson) +
-  theme(legend.position = "bottom")
-
-pp_check(canada_death_negbinomial) +
-  theme(legend.position = "bottom")
-```
-
-@fig-stanareyouokay-1 is a trace plot for the Poisson regression model.
+@fig-stanareyouokay-a-1 is a trace plot for the Poisson regression model.
 It shows a horizontal, dense band of samples without any systematic patterns, drifts, or long periods of stagnation.
@@
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows an Rhat value close to 1.  This suggests that the chains have converged to the target distribution.
+@fig-stanareyouokay-a-2 is a Rhat plot. It shows an Rhat value falling in the range 1-1.1.  This suggests that the chains have converged to the target distribution in the Poisson model.
 
@@
 This again suggests that the chains have converged to the target distribution in the negative regression model.
+The Rhat value is closer to 1 compared to in the Poisson model showing that the negative binomial regression model fits the data better.
 

### 2024-03-17 scripts/01-download_data.R

@@ -8,8 +8,9 @@
 # Any other information needed? N.A.
 
 #### Workspace setup ####
-
+install.packages("arrow")
 library(tidyverse)
+library(arrow)
 
 #### Download data ####
 raw_data <- 
@@ -17,6 +18,6 @@ raw_data <-
     file = "https://www150.statcan.gc.ca/t1/tbl1/en/dtl!downloadDbLoadingData-nonTraduit.action?pid=1310039401&latestN=0&startDate=20010101&endDate=20220101&csvLocale=en&selectedMembers=%5B%5B1%5D%2C%5B1%5D%2C%5B1%5D%2C%5B%5D%2C%5B1%2C2%5D%5D&checkedLevels=3D1%2C3D2")
 
 #### Save data from the Statistics Canada website  ####
-write_csv(raw_data, "inputs/data/raw_data.csv") 
+write_parquet(raw_data, "inputs/data/raw_data.parquet") 
 
          

### 2024-03-17 scripts/02-data_cleaning.R

@@ -12,8 +12,7 @@ library(janitor)
 library(dplyr)
 
 ### Clean the data ###
-raw_data <- read.csv("inputs/data/raw_data.csv")
-dim(raw_data)
+raw_data <- read_parquet("inputs/data/raw_data.parquet")
 
 # # Select the columns of interest
 col_name <- names(raw_data)
@@ -37,7 +36,6 @@ combined_data <- full_join(number_of_deaths, ranking_of_death, by = c("Year", "C
 # Filter out rows containing yearly totals
 cleaned_data <- filter(combined_data, !grepl("\\[A00-Y89\\]", Cause)) %>%
   mutate(Cause = sub("\\[.*", "", Cause))
-view(cleaned_data)
 
 #### Save data ####
-write_csv(cleaned_data, "outputs/data/cleaned_data.csv")
+write_parquet(cleaned_data, "outputs/data/cleaned_data.parquet")

### 2024-03-17 scripts/04-model.R

@@ -9,10 +9,12 @@
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(arrow)
 
 #### Read data ####
+cleaned_data <- read_parquet("outputs/data/cleaned_data.parquet")
 analysis_data <- cleaned_data |>
-  filter(Cause %in% canada_top_five)
+  filter(Cause %in% top_five_causes)
 
 ### Model data ####
 

### 2024-03-17 outputs/paper/paper.qmd
Updated the llm folder

@@
 thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
-date: yesterday
+date: today
 date-format: long
@@
 library(broom.mixed)
+library(loo)
 
@@
 This study investigates mortality in Canada from 2001 to 2022, utilizing the Canadian Vital Statistics Death Database by Statistics Canada.
-The focus is on identifying the risk factors associated with each cause of death.
+The estimand of the study is the relative risk factors for death associated with each cause of death.
 We employed a Bayesian approach, using both the Poisson regression model and the Negative Binomial regression model, to analyze the top five leading causes of mortality in Canada.
@@
 The Negative Binomial regression model addresses this by accounting for over-dispersion, offering a more flexible approach for such data.
+Through this study we also would like to compare the performance of the two models for the dataset we are interested in.
 
@@
 
-```{r}
-#| label: fig-2
-#| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
-#| echo: false
-#| warning: false
-#| message: false
-
-# cleaned_data |>
-#   filter(Year == 2021) |>
-#   mutate(Percentage = Death / sum(Death) * 100) |>
-#   ggplot(top_five, aes(x="", y= Death, fill= Cause)) +
-#   geom_bar(width = 1, stat = "identity") +
-#   coord_polar("y") +
-#   theme_void() +
-#   labs(fill="Cause of Death")
-
-```
-
 # Model {#sec-model}
@@
 ```{r}
-#|message= false
-#|warning= false
-#|include= false
+#| echo: false
+#| warning: false
+#| message: false
 
@@
   "CauseMalignant neoplasms" = "Malignant neoplasms",
-  "CauseCOVID-19" = "COVID-19"  # Ensure this matches the term exactly as it appears in your model output
+  "CauseCOVID-19" = "COVID-19"  
 )
 
-# Apply the corrected mapping to rename terms for the Poisson model
 tidy_poisson$term <- coef_short_names_corrected[tidy_poisson$term]
-# Handling any terms that did not match by retaining their original names
+
 tidy_poisson$term[is.na(tidy_poisson$term)] <- names(coef_short_names_corrected)[is.na(tidy_poisson$term)]
 
-# Apply the corrected mapping to rename terms for the Negative Binomial model
 tidy_negbinomial$term <- coef_short_names_corrected[tidy_negbinomial$term]
-# Handling any terms that did not match by retaining their original names
 tidy_negbinomial$term[is.na(tidy_negbinomial$term)] <- names(coef_short_names_corrected)[is.na(tidy_negbinomial$term)]
 
-# Combine the data for comparison
 comparison <- bind_rows(
@@
 
-# Assuming 'comparison' is your final data frame from previous steps
 comparison_table <- comparison %>%
@@
   )
+
+# LOO for both models
+loo_poisson <- loo(canada_death_poisson)
+loo_negbinomial <- loo(canada_death_negbinomial)
+
+# Assuming loo_poisson and loo_negbinomial have been calculated as shown:
+loo_poisson <- loo(canada_death_poisson, cores = 2)
+loo_negbinomial <- loo(canada_death_negbinomial, cores = 2)
+
+# Accessing LOOIC and ELPD correctly
+looic_poisson <- loo_poisson$looic
+elpd_loo_poisson <- loo_poisson$pointwise[, "elpd_loo"]
+elpd_se_poisson <- sqrt(var(elpd_loo_poisson))
+
+looic_negbinomial <- loo_negbinomial$looic
+elpd_loo_negbinomial <- loo_negbinomial$pointwise[, "elpd_loo"]
+elpd_se_negbinomial <- sqrt(var(elpd_loo_negbinomial))
+
+model_summary <- data.frame(
+  Metric = c("LOOIC", "ELPD", "ELPD SE"),
+  Poisson = c(looic_poisson, sum(elpd_loo_poisson), elpd_se_poisson),
+  Negative_Binomial = c(looic_negbinomial, sum(elpd_loo_negbinomial), elpd_se_negbinomial)
+)
 ```
@@
 comparison_table
+
 ```
@@
 Similarly, malignant neoplasms have a coefficient of approximately 1.805, indicating an expected count more than `r round(exp(1.805)-1, 2)` times higher than Accidents.
-Conversely, chronic lower respiratory diseases show a modest decrease in mortality compared to Accidents, with a coefficient of -0.070 in the Poisson model, indicating about `r round(1-exp(-0.070),2)` fewer deaths.
+Chronic lower respiratory diseases show a modest increase in mortality compared to Accidents, with a coefficient of 0.320 in the Poisson model and 0.322 in the negative binomial model, indicating about `r round(exp(0.320)-1,2)` times more deaths than Accidents.
+Lastly, Cerebrovascular diseases has a coefficient of 0.141, showing `r round(exp(0.141)-1,2)` times more death counts compare to Accidents although the impact is also modest.
 
-Table shows that among the top five leading causes of death in Canada, diseases of the heart with a coefficient of 1.454 and malignant neoplasms with a coefficient of approximately 1.805 significantly indicates a strong positive association in both models.
-The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model.
-Exponentiating the coefficient 1.454 to obtain suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries).
-Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is ` times higher Accidents (unintentional injuries).
-Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about  less.
+@tbl-4 displayed the values of the parameters in both models.The LOOIC value (Leave-One-Out Cross-Validation Information Criterion) is useful in comparing model fitness.
+A lower LOOIC value suggests a model that is expected to make more accurate predictions for new, unseen data.
+This is because LOOIC penalizes models for having too many parameters (complexity), helping to avoid overfitting.
+The Poisson model has a LOOIC value of 33604.96 whereas the negative binomial model has a LOOIC value of 1787.36 much lower than the Poisson model.
+This indicates a better fit of the data by the negative binomial model.
+
+```{r}
+#| label: tbl-4
+#| tbl-cap: "Model summary of Poisson model vs. negative binomial model"
+#| echo: false
+#| warning: false
+#| message: false
+
+model_summary %>%
+  gt() %>%
+  tab_header(
+    title = "Model Summary: Poisson vs. Negative Binomial"
+  ) %>%
+  cols_label(
+    Metric = "Metric",
+    Poisson = "Poisson",
+    Negative_Binomial = "Negative Binomial"
+  ) %>%
+  fmt_number(
+    columns = vars(Poisson, Negative_Binomial),
+    decimals = 2
+  )
+```
 
@@
 The negative binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794.
-The negative binomial model's better fit is further supported by a higher ELPD value compared to the Poisson model as shown in @tbl-4.
+The negative binomial model's better fit is further supported by a higher ELPD value compared to the Poisson model as shown in @tbl-5.
 Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.
+\newpage
 
 ```{r}
-#| label: tbl-4
+#| label: tbl-5
 #| tbl-cap: "Cross validation comparision between Poisson model and negative binomial model "
@@
 
-(loo_compare(poisson, neg_binomial))
+comparison_result <- loo_compare(poisson, neg_binomial)
+comparison_df <- as.data.frame(comparison_result)
+
+gt_table <- gt(comparison_df) %>%
+  tab_header(
+    title = "Cross-Validation Comparison",
+    subtitle = "Poisson model vs. Negative Binomial model"
+  )
+
+gt_table
 
@@
 
+The negative binomial regression model is well-suited for datasets exhibiting overdispersion, which is evidenced in our dataset by various model parameters.
+This overdispersion signifies that the variance of the dependent variable is larger than would be expected under a Poisson distribution, thus making the negative binomial model a more appropriate choice due to its ability to account for this variability.
+
 ## Ethical implications
@@
 
-Future studies could aim to address these weaknesses by incorporating spatial and demographic variables to account for clustering effects.
-Future studies can broaden the scope to include more regions, socioeconomical variables, and their interactions.
+Future studies could aim to address these weaknesses by incorporating spatial and demographic variables to account for clustering effects.For instance, additional provinces could be included in the investigation to compare and elucidate factors impacting regional mortality.
+Future studies can broaden the scope to include more socioeconomical variables, such as diet, lifestyles, environmental pollutants and their interactions.
+Canada is also home to omany centenarians in the world [@statcan]; conversely, we could compare mortality and the phenomenon of longevity.
 Longitudinal studies could offer insights into mortality trends and the long-term effects of environmental factors like climate change.
@@
 The "Characteristics" variable is converted into a new variable "Death" indicating the number of death attributed to the cause and a new "Rank" variable showing the rank of the cause in a given year.
-@tbl-5 presents a sample of the cleaned dataset.
+@tbl-6 presents a sample of the cleaned dataset.
 
 ```{r}
-#| label: tbl-5
+#| label: tbl-6
 #| tbl-cap: "A sample of the cleaned dataset of number of death attributed to causes of death in Canada, 2001-2022"

### 2024-03-17 scripts/00-simulate_data.R
Updated simulation and test

@@ -1,19 +1,38 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates dataset 
+# Author: Yuchao Niu
+# Date: 16 March 2024
+# Contact: yc.niu@utornoto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: N.A.
+# Any other information needed? N.A.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(dplyr)
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(123) # Ensure reproducibility
+
+# Create a dataset
+num_observations <- 120
+years <- sample(2001:2023, num_observations, replace=TRUE)
+causes_of_death <- c('heart disease', 'kidney disease', 'cancer', 'lung disease', 'neuro', 'skin', 'skeletomuscular')
+death_causes <- sample(causes_of_death, num_observations, replace=TRUE)
+number_of_deaths <- sample(100:1000, num_observations, replace=TRUE)
+
+sim_data <- data.frame(Year=years, 'Cause of Death'=death_causes, 'Number of Deaths'=number_of_deaths)
+
+# Calculate the total number of deaths by cause
+death_counts_by_cause <- aggregate(dataset$`Number of Deaths`, by=list(Cause=dataset$`Cause of Death`), FUN=sum)
+names(death_counts_by_cause) <- c('Cause of Death', 'Total Deaths')
+
+# Rank the causes by the number of deaths
+death_counts_by_cause$Rank <- rank(-death_counts_by_cause$`Total Deaths`, ties.method='min')
+
+# Merge the ranking back to the original dataset
+dataset_with_rank <- merge(dataset, death_counts_by_cause[,c('Cause of Death', 'Rank')], by='Cause of Death')
 
 
 

### 2024-03-17 scripts/03-test_data.R

@@ -1,15 +1,84 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests Mortality Dataset
+# Author: Yc.niu@utoro
+# Contact: yc.niu@utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: N.A.
+# Any other information needed? N.A.
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(arrow)
 
-#### Test data ####
+#### Test Data ####
+# read cleaned data #
+cleaned_data <- read_parquet("outputs/data/cleaned_data.parquet")
+
+# Test for Valid Years Range
+if(all(cleaned_data$Year >= 2001 & cleaned_data$Year <= 2023)) {
+  print("Test Passed: All years are within the 2001 to 2023 range")
+} else {
+  print("Test Failed: Some years are outside the 2001 to 2023 range")
+}
+
+# Test for Non-negative Number of Deaths
+if (all(cleaned_data$Death >= 0, na.rm = TRUE)) {
+  print("Test Passed: Number of Deaths is non-negative")
+} else {
+  print("Test Failed: There are negative values in Number of Deaths")
+}
+
+# Test for No Missing Values
+if(sum(is.na(cleaned_data)) == 0) {
+  print("Test Passed: No missing values in the dataset")
+} else {
+  print("Test Failed: There are missing values in the dataset")
+}
+
+# Test for Unique Ranks per year
+cleaned_data %>%
+  group_by(Year, Cause) %>%
+  summarise(is_unique_rank = n_distinct(Ranking) == n()) %>%
+  ungroup() %>%
+  filter(is_unique_rank == FALSE) -> non_unique_rank_cases
+
+if (nrow(non_unique_rank_cases) == 0) {
+  print("Test Passed: Ranks are unique for each cause of death per year")
+} else {
+  print("Test Failed: Ranks are not unique for each cause of death per year")
+}
+
+# Test for No Duplicate Observations
+if(nrow(cleaned_data) == nrow(unique(cleaned_data))) {
+  print("Test Passed: No duplicate observations")
+} else {
[…]

### 2024-03-17 scripts/00-simulate_data.R
final revision

@@ -1,38 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates dataset 
-# Author: Yuchao Niu
-# Date: 16 March 2024
-# Contact: yc.niu@utornoto.ca
-# License: MIT
-# Pre-requisites: N.A.
-# Any other information needed? N.A.
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-
-#### Simulate data ####
-set.seed(123) # Ensure reproducibility
-
-# Create a dataset
-num_observations <- 120
-years <- sample(2001:2023, num_observations, replace=TRUE)
-causes_of_death <- c('heart disease', 'kidney disease', 'cancer', 'lung disease', 'neuro', 'skin', 'skeletomuscular')
-death_causes <- sample(causes_of_death, num_observations, replace=TRUE)
-number_of_deaths <- sample(100:1000, num_observations, replace=TRUE)
-
-sim_data <- data.frame(Year=years, 'Cause of Death'=death_causes, 'Number of Deaths'=number_of_deaths)
-
-# Calculate the total number of deaths by cause
-death_counts_by_cause <- aggregate(dataset$`Number of Deaths`, by=list(Cause=dataset$`Cause of Death`), FUN=sum)
-names(death_counts_by_cause) <- c('Cause of Death', 'Total Deaths')
-
-# Rank the causes by the number of deaths
-death_counts_by_cause$Rank <- rank(-death_counts_by_cause$`Total Deaths`, ties.method='min')
-
-# Merge the ranking back to the original dataset
-dataset_with_rank <- merge(dataset, death_counts_by_cause[,c('Cause of Death', 'Rank')], by='Cause of Death')
-
-
-

### 2024-03-17 scripts/03-test_data.R

@@ -1,84 +1,0 @@
-#### Preamble ####
-# Purpose: Tests Mortality Dataset
-# Author: Yc.niu@utoro
-# Contact: yc.niu@utoronto.ca
-# License: MIT
-# Pre-requisites: N.A.
-# Any other information needed? N.A.
-
-#### Workspace setup ####
-library(tidyverse)
-library(arrow)
-
-#### Test Data ####
-# read cleaned data #
-cleaned_data <- read_parquet("outputs/data/cleaned_data.parquet")
-
-# Test for Valid Years Range
-if(all(cleaned_data$Year >= 2001 & cleaned_data$Year <= 2023)) {
-  print("Test Passed: All years are within the 2001 to 2023 range")
-} else {
-  print("Test Failed: Some years are outside the 2001 to 2023 range")
-}
-
-# Test for Non-negative Number of Deaths
-if (all(cleaned_data$Death >= 0, na.rm = TRUE)) {
-  print("Test Passed: Number of Deaths is non-negative")
-} else {
-  print("Test Failed: There are negative values in Number of Deaths")
-}
-
-# Test for No Missing Values
-if(sum(is.na(cleaned_data)) == 0) {
-  print("Test Passed: No missing values in the dataset")
-} else {
-  print("Test Failed: There are missing values in the dataset")
-}
-
-# Test for Unique Ranks per year
-cleaned_data %>%
-  group_by(Year, Cause) %>%
-  summarise(is_unique_rank = n_distinct(Ranking) == n()) %>%
-  ungroup() %>%
-  filter(is_unique_rank == FALSE) -> non_unique_rank_cases
-
-if (nrow(non_unique_rank_cases) == 0) {
-  print("Test Passed: Ranks are unique for each cause of death per year")
-} else {
-  print("Test Failed: Ranks are not unique for each cause of death per year")
-}
-
-# Test for No Duplicate Observations
-if(nrow(cleaned_data) == nrow(unique(cleaned_data))) {
-  print("Test Passed: No duplicate observations")
-} else {
-  print("Test Failed: There are duplicate observations")
-}
-
-# Test for Correct Data Types
-
-## Test for 'Death' column being numeric
-if (class(cleaned_data$Death) == "numeric") {
-  message("Test Passed: 'Death' column is numeric.")
-} else {
-  message("Test Failed: 'Death' column is not numeric.")
-}
-
-## Test for 'Ranking' column being numeric
-if (class(cleaned_data$Ranking) == "integer") {
[…]
